A simplified Plackett-Luce likelihood function for rank orders

### 2024-03-12 inst/hankin-ejs.tex
follow instructions on ejs website

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ cloned competitors, each of strength $b=1-a$.  An observation is
indexed by~$r$, the number of strength $b$ clones finishing ahead
of~$a$, and because we have $n+1$ competitors $0\leqslant r\leqslant
n$.  The initial field strength is $a+nb=n+(n-1)a$.  The Plackett-Luce
likelihood function~\cite{plackett1975,luce1959} for this situation
would be

@@

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+ K + \mathcal{O}\left(n^{-2}\right)
\end{equation}

@@

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  I present some numerical results below.

Previous related work presented the case $n=2$ [in current notation].
Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
between $r$ and $n-r$.  For $r=3$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.64$.

@@

One issue encountered in educational research is that of
confidentiality of examination scores, and individual results are
generally regarded as sensitive and private information.
Nevertheless, it is common a student to possess some informative
observations: the total enrollment in a specific class and their
individual ranking within that class;  table~\ref{educationtable}
shows an apocryphal dataset.   What could a student infer from such
data?

@@

One natural question would be to wonder whether the student was better
at pure or applied mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as a
diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion is met, and we are justified in concluding that the
student is indeed more competent at applied than pure mathematics.
Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
region of its asymptotic $\chi^2_1$ distribution, corresponding to a
$p$-value of about $0.028$.

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ cloned competitors, each of strength $b=1-a$.  An observation is
indexed by~$r$, the number of strength $b$ clones finishing ahead
of~$a$, and because we have $n+1$ competitors %%%%%%%%%%%%%%%%%%%%%%
$0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=n+(n-1)a$.  The Plackett-Luce likelihood
function~\cite{luce1959,plackett1975} for this situation would be

@@

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+ K + \mathcal{O}\left(n^{-2}\right).
\end{equation}

@@

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  Previous related work presented the case $n=2$ [in
  current notation].  Considering the case $n=7$ gives us $7+1=8$
distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
unanticipated asymmetry between $r$ and $n-r$.  For $r=3$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.64$.

@@

One issue encountered in educational research is that of
confidentiality of examination scores, and individual results are
generally regarded as sensitive and private information.
Nevertheless, it is common for a student to possess some informative
observations: the total enrollment in a specific class and their
individual ranking within that class; table~\ref{educationtable} shows
an apocryphal dataset.  What could a student infer from such data?

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the
diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion is met, and we are justified in concluding that the
student is indeed more competent at applied than pure mathematics.
Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
region of its asymptotic $\chi^2_1$ distribution, corresponding to a
$p$-value of about $0.028$.
